---
title: 'Reproducible Research: Peer Assessment 1'
author: "Alessandra Coletti"
date: "10 luglio 2016"
output: html_document
---

## Loading and preprocessing the data
### Load the data
```{r}
setwd("C:/Users/IBM_ADMIN/Desktop/personale/Coursera/Data Science Specialization/5 - Reproducible Research/Week 2")
activity <- read.csv("activity.csv")
```
####  See what our data frame is all about
```{r}
names(activity)
```
```{r}
str(activity)
```
```{r}
head(activity,10)
```
### Process/transform the data into a format suitable for your analysis
###### subset data frame to values without NA for later use
```{r}
without_na <- activity[complete.cases(activity),]
```
## What is mean total number of steps taken per day?
### Make a histogram of the total number of steps taken each day
##### total number of steps taken per day
```{r}
total <- aggregate(steps ~ date, without_na, sum)
```
##### add descriptive variable names
```{r}
names(total)[2] <- "sum_steps"
```
##### check out new data frame
```{r}
head(total, 5)
```
##### plot histogram
```{r,echo=TRUE}
hist(
        total$sum_steps,
        col = "blue",
        main = "Histogram of the Total Number of Steps Taken Each Day",
        xlab = "Total Number of Steps Taken Each Day",
        breaks = 20
)
```

### Calculate and report the mean and median total number of steps taken per day

##### mean
```{r}
mean(total$sum_steps)
```
##### median
```{r}
median(total$sum_steps)
```
## What is the average daily activity pattern?
### Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
##### the average number of steps taken, averaged across all days for each 5-minute interval
```{r}
interval <- aggregate(steps ~ interval, without_na, mean)
```
##### add descriptive variable names
```{r}
names(interval)[2] <- "mean_steps"
```
##### check out new data frame
```{r}
head(interval, 5)
```
##### format plot margins to accommodate long text labels
```{r}
par(mai = c(1.2,1.5,1,1))
```
##### plot time series
```{r,echo=TRUE}
plot(
        x = interval$interval,
        y = interval$mean_steps,
        type = "l",
        main = "Time Series Plot of the 5-Minute Interval\n and the Average Number of Steps Taken, Averaged Across All Days",
        xlab = "5-Minute Interval",
        ylab = "Average Number of Steps Taken,\n Averaged Across All Days"
)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
interval[interval$mean_steps==max(interval$mean_steps),]
```
## Imputing missing values
### Calculate and report the total number of missing values in the dataset
```{r}
nrow(activity[is.na(activity$steps),])
```
### Devise a strategy for filling in all of the missing values in the dataset
#### I will use the mean for the 5-minute interval to populate NA values for a given interval
### Create a new dataset that is equal to the original dataset but with the missing data filled in
##### merge original activity data frame with interval data frame
```{r}
newactivity <- merge(activity, interval, by = 'interval', all.y = F)
```
##### merge NA values with averages rounding up for integers
```{r}
newactivity$steps[is.na(newactivity$steps)] <- as.integer(
        round(newactivity$mean_steps[is.na(newactivity$steps)]))
```
##### drop and reorder columns to match original activity data frame
```{r}
keeps <- names(activity)
newactivity <- newactivity[keeps]
```
### Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day
##### total number of steps taken per day
```{r}
newtotal <- aggregate(steps ~ date, newactivity, sum)
```
##### add descriptive variable names
```{r}
names(newtotal)[2] <- "sum_steps"
```
##### check out new data frame
```{r}
head(newtotal, 5)
```
##### plot histogram
```{r,echo=TRUE}
hist(
        newtotal$sum_steps,
        col = "blue",
        main = "Histogram of the Total Number of Steps Taken Each Day\nPart Deux",
        xlab = "Total Number of Steps Taken Each Day Part Deux",
        breaks = 20
)
```

##### mean
```{r}
mean(newtotal$sum_steps)
```
##### median
```{r}
median(newtotal$sum_steps)
```
### Do these values differ from the estimates from the first part of the assignment?
#### They do differ, but ever so slightly:
#### mean(total) = 10766.19, while mean(newtotal) = 10765.64 - rounding produces the same value
#### median(total) = 10765, while median(newtotal) = 10762 - 3 step difference

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
#### This seems to highly depend on how you impute the missing data. Since I used the average for a given interval, there was practically no difference because we basically pulled the averages closer to the inserted average value.

## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day
##### create new data frame
```{r}
newnewactivity <- newactivity
```
##### set up logical/test vector
```{r}
weekend <- weekdays(as.Date(newnewactivity$date)) %in% c("sabato", "domenica")
```
##### Fill in weekday column
```{r}
newnewactivity$daytype <- "weekday"
```
##### replace "weekday" with "weekend" where day == Sat/Sun
```{r}
newnewactivity$daytype[weekend == TRUE] <- "weekend"
```
##### convert new character column to factor
```{r}
newnewactivity$daytype <- as.factor(newnewactivity$daytype)
```
##### Check out new data frame
```{r}
str(newnewactivity)
```
```{r}
head(newnewactivity, 5)
```
##### double check
```{r}
weekdays(as.Date(newnewactivity$date[3]))
```

### Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
##### the average number of steps taken, averaged across all days for each 5-minute interval
```{r}
newinterval <- aggregate(steps ~ interval + daytype, newnewactivity, mean)
```
##### add descriptive variable names
```{r}
names(newinterval)[3] <- "mean_steps"
```
##### check out new data frame
```{r}
head(newinterval, 5)
```
##### plot time series
```{r,echo=TRUE}
library(lattice)
xyplot(
        mean_steps ~ interval | daytype,
        newinterval,
        type = "l",
        layout = c(1,2),
        main = "Time Series Plot of the 5-Minute Interval\nand the Average Number of Steps Taken,\nAveraged Across All Weekday Days or Weekend Days",
        xlab = "5-Minute Interval",
        ylab = "Average Number of Steps Taken"
)
```